Commutative Character Sheaves, with an application to generic types for supercuspidal representations

### 2015-08-31 followup.tex
Sorted out descent to the abelianization, I think! Sketched the argument.
Will add detail, later.

Let $G$ be a smooth group scheme.
Let $G\der \hookrightarrow G$ denote the closed subgroup generated by the commutator morphism $c: G\times G \to G$ given by $c(x,y) = xyx^{-1}y^{-1}$; see \cite{SGA3}*{{VI}$_B$, 7.2}.

\begin{lemma}
The quotient $q : G \to G\ab$ is a smooth $G\der$-torsor, and the coequalizer of $m, p : G\der \times G \to G$.
\end{lemma}

The quotient $q : G \to G\ab$ is the coequalizer of left $G\der$-multiplication $n : G\der\times G \to G$ and projection $p : G\der\times G \to G$.
\[
\begin{tikzcd}
G\der\times G
 \arrow{r}[shift left]{m}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G\ab
\end{tikzcd}
\]
The morphism $q$ is surjective and smooth of relative dimension $\dim G\der$.

@@

\begin{proof}
It follows from the lemma that descent theory that $q$ is an 
effective descent morphism for the fibred category of commutative character sheaves on smooth group schemes; in other words, the canonical functor $\CCS(G) \to \CCS_{G\der}(G)$ is an equivalence. 
\todo{$\CCS \to \Sch{\Fq}$ is a fibred category. Is it also a stack?}

Consider the morphism $c:  G \times G \to G\der$ given by $c(g,h)= ghg^{-1}h^{-1}$.
This morphism is smooth (it is a composition of smooth maps, since $m : G \times G \to G$ is smooth) but

Let $j : C \hookrightarrow G$ be the inclusion.  
Then $j\circ c = m\circ (m \times (\iota\circ m\op)) : G \times G \to G$ is $(g,h) \mapsto gh g^{-1}h^{-1}$ while $m\op\circ (m \times (\iota\circ m\op))= e : G \times G \to G$ is $(g,h) \mapsto 1_G$. Thus,
\begin{align*}
c^* (\cs{L}\vert_C)
&=  (j\circ c)^*\cs{L} \\
&= (m\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= (m \times (\iota\circ m\op))^* m^* \cs{L} \\
&= (m \times (\iota\circ m\op))^* m\op^* \cs{L} \\
&= (m\op\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= e^* \cs{L}\\
&= (\EE)_{G\times G}\\
&= c^* ((\EE)_C).
\end{align*}
Since $c : G \times G \to C$ is faithfully flat, the functor $c^*$ is conservative.
\todo{I'm working on this   . . . \\ -CC}
Since $\cs{L} \to (\EE)_G$ induces $c^* (\cs{L}\vert_C) \iso c^* ((\EE)_C)$, this implies $\cs{L}\vert_C \iso (\EE)_C$.

Let $\CCS_{G\der}(G)$ be the category of commutative character sheaves $\cs{L}$ on $G$ equipped with an isomorphism $n^*\cs{L} \to p^*\cs{L}$ satisfying canonical coherence conditions.
Then $q^* : \CCS(G\ab) \to \CCS(G)$ factors through $\CCS_{G\der}(G)$. 
\[
\begin{tikzcd}
\CCS(G\der \times G)
& \arrow{l}[shift left,swap]{n^*}
  \arrow{l}[shift right]{p^*}
\CCS(G) 
&&\arrow{ll}[swap]{q^*} 
\CCS(G\ab) \arrow{dl}{\text{comparison}}\\
&& \arrow{ul}{\text{forget}} \CCS_{G\der}(G) & 
\end{tikzcd}
\]
To see this, let $\cs{E}$ be a commutative character sheaf on $G\ab$ and observe that $q^* \cs{E}$ is equipped with a canonical isomorphism $n^*(q^* \cs{E}) \to p^*(q^* \cs{E})$ given by
\[
n^* (q^*\cs{E}) \iso (q\circ n)^* \cs{E} = (q\circ p)^* \cs{E} \iso p^* (q^*\cs{E}).
\]
Moreover, since $q : G \to G\ab$ is a smooth principal $G\der$-bundle and $G\der$ is geometrically connected,
the comparison functor is an equivalence.
%(In other language, $q^*$ is a monadic functor.)
Indeed, it is easy to see that $\CCS(G\ab) \to \CCS_{G\der}(G)$ is full and faithful; to see that is it also essentially surjective, one begins with $\cs{L} \in \CCS(G)$ and $\nu : n^*\cs{L} \to p^*\cs{L}$ and builds $\cs{E}\in \CCS(G\ab)$ Zariski-locally by pull-back along local trivializations $U \to G\times_{G\ab} U$ of $q : G\to G\ab$ compatible with the $C$-action, glueing using isomorphisms inherited from the isomorphism $\nu$, and finally verifying that the result is a local system with the auxilliary data defining a commutative character sheaf. See appendix for details.

To see that the forgetful functor $\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence we observe that
every commutative character sheaf $\cs{L}$ on $G$ is equivariant for the left translation action of $C$ on $G$, as follows.
Let $j : C \hookrightarrow G$ be the kernel of $q : G \to G\ab$ and let $m : G\times G \to G$ be multiplication for $G$. 
Then $n = m \circ (j \times \id)$, so
\begin{align*}
n^* \cs{L} 
&= (m \circ (j \times \id))^* \cs{L}\\
&= (j \times \id))^* m^*\cs{L}\\
&\iso (j \times \id))^* (\cs{L} \boxtimes \cs{L})\\
&= (j^*\cs{L}) \boxtimes \cs{L}\\
&= (\cs{L}\vert_C) \boxtimes \cs{L}\\
&= (\EE)_C \boxtimes \cs{L}\\
&= p^* \cs{L}.
\end{align*}

\end{proof}

\begin{remark}\label{rem:descent}
\end{remark}

>>>>

Let $G$ be a smooth group scheme.
Let $j: G\der \hookrightarrow G$ denote the closed subgroup generated by the commutator morphism $c: G\times G \to G$ given by $c(x,y) = xyx^{-1}y^{-1}$; see \cite{SGA3}*{{VI}$_B$, 7.2}.
Let $n : G\der \times G \to G$ be the restriction of $m: G\times G \to G$.

\begin{lemma}\label{lem:Lder}
For every $\cs{L}$ on $G$, $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$.
\end{lemma}

\begin{proof}
By construction, $c : G \times G \to G$ is schematically dense (dominant morphism) for $j : G\der \hookrightarrow G$. 
Note that $j\circ c = m\circ (m \times (\iota\circ m\op)) : G \times G \to G$ is $(g,h) \mapsto gh g^{-1}h^{-1}$ while $m\op\circ (m \times (\iota\circ m\op))= e : G \times G \to G$ is $(g,h) \mapsto 1_G$. Thus,
\begin{align*}
c^* (\cs{L}\vert_{G\der})
&=  (j\circ c)^*\cs{L} \\
&= (m\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= (m \times (\iota\circ m\op))^* m^* \cs{L} \\
&\iso (m \times (\iota\circ m\op))^* m\op^* \cs{L} \\
&= (m\op\circ (m \times (\iota\circ m\op)))^* \cs{L} \\
&= e^* \cs{L}\\
&= (\EE)_{G\times G}.
\end{align*}
Since $c : G \times G \to G\der$ is schematically dense, the group homomorphism $\pi_1(G\times G, {\bar y}) \to \pi_1(G\der, {\bar x})$ is surjective, for all geometric points with $c({\bar y}) = {\bar x}$. It follows that $c^* (\cs{L}\vert_{G\der}) \iso (\EE)_{G\times G}$ implies $\cs{L}\vert_{G\der} \iso (\EE)_{G\der}$. 
\end{proof}

\begin{lemma}\label{lem:smooth_coeq}
The quotient $q : G \to G\ab\ceq G/G\der$ is a smooth $G\der$-torsor and is the coequalizer of $n : G\der \times G \to G$ with projection $p: G\der \times G \to G$. 
\[
\begin{tikzcd}
G\der\times G
 \arrow{r}[shift left]{n}
  \arrow{r}[shift right,swap]{p}
&
G 
\arrow{r}{q}
& 
G\ab
\end{tikzcd}
\]
\end{lemma}

\begin{proposition}\label{prop:Gder-equivariance}
$\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence.
\end{proposition}

\begin{proof}
It follows from Lemma~\ref{lem:Lder} that the forgetful functor $\CCS_{G\der}(G) \to \CCS(G)$ is an equivalence. To see this, we observe that
every commutative character sheaf $\cs{L}$ on $G$ is equivariant for the left translation action of $C$ on $G$, as follows.
Let $j : C \hookrightarrow G$ be the kernel of $q : G \to G\ab$ and let $m : G\times G \to G$ be multiplication for $G$. 
Then $n = m \circ (j \times \id)$, so
\begin{align*}
n^* \cs{L} 
&= (m \circ (j \times \id))^* \cs{L}\\
&= (j \times \id))^* m^*\cs{L}\\
&\iso (j \times \id))^* (\cs{L} \boxtimes \cs{L})\\
&= (j^*\cs{L}) \boxtimes \cs{L}\\
&= (\cs{L}\vert_C) \boxtimes \cs{L}\\
&= (\EE)_C \boxtimes \cs{L}\\
&= p^* \cs{L}.
\end{align*}
Thus, $\cs{L}$ is $G\der$-equivariant.
\end{proof}

@@

\begin{proof}
It follows from Lemma~\ref{lem:smooth_coeq} that $q$ is an 
effective descent morphism for the fibred category of commutative character sheaves on smooth group schemes; in other words, the canonical functor $\CCS(G) \to \CCS_{G\der}(G)$ is an equivalence. 
\todo{$\CCS \to \Sch{\Fq}$ is a fibred category. Is it also a stack?}
See appendix for details.
%
Combining this with Proposition~\ref{prop:Gder-equivariant}, it follows that $q^* : \CCS(G\ab)\to \CCS(G)$ is an equivalence.
\end{proof}
